---
title: "rbiips demo"
author: 'Adrien Todeschini, François Caron'
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{rbiips demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

This is a quick demo of `rbiips` main functions.

Disclaimer: for speed of execution, we use very few particles and MCMC iterations in this vignette.
As such, the results are not representative of the perforance of SMC and particle MCMC methods.


```{r}
set.seed(0)
library(rbiips)
```

# Add custom functions and distributions to BUGS language

Add custom function `f`

```{r}
f_dim <- function(x_dim, t_dim) {
  # Check dimensions of the input and return dimension of the output of function f
  stopifnot(prod(x_dim) == 1, prod(t_dim) == 1)
  x_dim
}
f_eval <- function(x, t) {
  # Evaluate function f
  0.5 * x + 25 * x/(1 + x^2) + 8 * cos(1.2 * t)
}
biips_add_function('f', 2, f_dim, f_eval)
```

Add custom sampling distribution `dMN`

```{r}
dMN_dim <- function(mu_dim, Sig_dim) {
  # Check dimensions of the input and return dimension of the output of
  # distribution dMN
  stopifnot(prod(mu_dim) == mu_dim[1], length(Sig_dim) == 2, mu_dim[1] == Sig_dim)
  mu_dim
}
dMN_sample <- function(mu, Sig) {
  # Draw a sample of distribution dMN
  mu + t(chol(Sig)) %*% rnorm(length(mu))
}
biips_add_distribution('dMN', 2, dMN_dim, dMN_sample)
```

# Compile model

```{r}
modelfile <- system.file('extdata', 'hmm_f.bug', package = 'rbiips')
stopifnot(nchar(modelfile) > 0)
cat(readLines(modelfile), sep = '\n')

tmax <- 10
data_ <- list(tmax = tmax, p = c(.5, .5), logtau_true = log(1), logtau = log(1))
model <- biips_model(modelfile, data_, sample_data = TRUE)
data_ = model$data()
```

# SMC algorithm

```{r}
n_part <- 100
out_smc <- biips_smc_samples(model, c('x', 'c[2:10]', 'x[5]', 'c[5]'), 
                             n_part, type = 'fs',
                             rs_thres = 0.5, rs_type = 'stratified')

biips_diagnosis(out_smc)
summ_smc <- biips_summary(out_smc, probs=c(0.025, 0.975))

par(mfrow = c(2, 2), bty='l')

matplot(do.call(cbind, summ_smc$x$f$quant), 
        type='n', xlab='t', ylab='x[t]')
lines(summ_smc$x$f$mean)
matlines(do.call(cbind, summ_smc$x$f$quant), col=1, lty=2)
lines(summ_smc$x$s$mean, col=2)
matlines(do.call(cbind, summ_smc$x$s$quant), col=2, lty=2)
lines(data_$x_true, col=3)
legend('topleft', 
       leg=c('SMC filtering estimate',
             'SMC smoothing estimate', 'true'),
       col=1:3, lty=1, bty='n')

matplot(c(1,tmax), c(0,4), type='n', xlab='t', ylab='c[t]==1', yaxt='n')
lines(2:tmax, 2+summ_smc[['c[2:10]']]$f$mode, type='S')
lines(2:tmax, .5+summ_smc[['c[2:10]']]$s$mode, type='S', col=2)
lines(-1+data_$c_true, type='S', col=3)
legend('topleft', 
       leg=c('SMC filtering mode',
             'SMC smoothing mode', 'true'),
       col=1:3, lty=1, bty='n')

plot(biips_density(out_smc[['x[5]']]))
abline(v=data_$x_true[5], col=3)

plot(biips_table(out_smc[['c[5]']]))
points(data_$c_true[5], 0, col=3, pch=17)
```

# PIMH algorithm

```{r}
n_part <- 50
obj_pimh <- biips_pimh_init(model, c('x', 'c[2:10]', 'x[5]', 'c[5]'))  # Initialize
out_pimh_burn <- biips_pimh_update(obj_pimh, 100, n_part)  # Burn-in
out_pimh <- biips_pimh_samples(obj_pimh, 100, n_part)  # Samples

summ_pimh <- biips_summary(out_pimh, probs=c(0.025, 0.975))

par(mfrow = c(2, 1), bty='l')

plot(c(out_pimh_burn$log_marg_like, out_pimh$log_marg_like),
     type = 'l', xlab = 'PIMH iteration', ylab = 'log p(y)')

plot(101:200, c(out_pimh[['x[5]']]),
     type = 'l', xlab = 'PIMH iteration', ylab = 'x[5]')
abline(h=data_$x_true[5], col = 3)

par(mfrow = c(2, 2), bty='l')

matplot(do.call(cbind, summ_pimh$x$quant), 
        type='n', xlab='t', ylab='x[t]')
lines(summ_pimh$x$mean)
matlines(do.call(cbind, summ_pimh$x$quant), col=1, lty=2)
lines(data_$x_true, col=3)
legend('topleft', 
       leg=c('PIMH estimate', 'true'),
       col=c(1,3), lty=1, bty='n')

matplot(c(1,tmax), c(0,2.5), type='n', xlab='t', ylab='c[t]==1', yaxt='n')
lines(2:tmax, .5+summ_pimh[['c[2:10]']]$mode, type='S')
lines(-1+data_$c_true, type='S', col=3)
legend('topleft', 
       leg=c('PIMH mode', 'true'),
       col=c(1,3), lty=1, bty='n')

plot(biips_density(out_pimh[['x[5]']]))
abline(v=data_$x_true[5], col=3)

plot(biips_table(out_pimh[['c[5]']]))
points(data_$c_true[5], 0, col=3, pch=17)
```

# SMC sensitivity analysis

```{r}
n_part <- 50
logtau_val <- -10:10
out_sens <- biips_smc_sensitivity(model, list(logtau = logtau_val), n_part)

par(mfrow = c(2,1), bty='l')
plot(logtau_val, out_sens$log_marg_like, type='l',
     xlab = 'logtau', ylab = 'log p(y | logtau)')
abline(v=data_$logtau_true, col=3)

plot(logtau_val, out_sens$log_marg_like_pen, type='l',
     xlab = 'logtau', ylab = 'log p(y, logtau)')
abline(v=data_$logtau_true, col=3)
```

# PMMH algorithm

```{r}
data_ <- list(tmax = 10, p = c(.5, .5), logtau_true = log(1))
model <- biips_model(modelfile, data_)
data_ <- model$data()

n_part <- 50
n_burn <- 100
n_iter <- 100
obj_pmmh <- biips_pmmh_init(model, 'logtau', 
                            latent_names = c('x', 'c[2:10]', 'x[5]', 'c[5]'),
                            inits = list(logtau = -1))  # Initialize
out_pmmh_burn <- biips_pmmh_update(obj_pmmh, n_burn, n_part)  # Burn-in
out_pmmh <- biips_pmmh_samples(obj_pmmh, n_iter, n_part, thin = 1)  # Samples

summ_pmmh <- biips_summary(out_pmmh, probs=c(.025, 0.975))

par(mfrow = c(2, 2), bty='l')

plot(c(out_pmmh_burn$log_marg_like_pen, out_pimh$log_marg_like_pen),
     type = 'l', xlab = 'PMMH iteration', ylab = 'log p(y, logtau)')

plot(n_burn+(1:n_iter), c(out_pmmh$logtau),
     type = 'l', xlab = 'PMMH iteration', ylab = 'logtau')
abline(h=data_$logtau_true, col = 3)

plot(biips_density(out_pmmh$logtau))
abline(v=data_$logtau_true, col=3)

biips_hist(out_pmmh$logtau)
abline(v=data_$logtau_true, col=3)

matplot(do.call(cbind, summ_pmmh$x$quant), 
        type='n', xlab='t', ylab='x[t]')
lines(summ_pmmh$x$mean)
matlines(do.call(cbind, summ_pmmh$x$quant), lty=2, col=1)
lines(data_$x_true, col=3)
legend('topleft', 
       leg=c('PMMH estimate', 'true'),
       col=c(1,3), lty=1, bty='n')

matplot(c(1,tmax), c(0,2.5), type='n', xlab='t', ylab='c[t]==1', yaxt='n')
lines(2:tmax, .5+summ_pmmh[['c[2:10]']]$mode, type='S')
lines(-1+data_$c_true, type='S', col=3)
legend('topleft', 
       leg=c('PMMH mode', 'true'),
       col=c(1,3), lty=1, bty='n')

plot(biips_density(out_pmmh[['x[5]']]))
abline(v=data_$x_true[5], col=3)

plot(biips_table(out_pmmh[['c[5]']]))
points(data_$c_true[5], 0, col=3, pch=17)
```
